Generalized Quasi-linear Approximation and Non-normality in Taylor-Couette Flow

### 2022-07-31 paper/paper.tex

\author{%%%% Author details
Jeffrey S. Oishi$^{1}$ and Morgan Baxter$^{1,2}$}

%%%%%%%%% Insert author address here
\address{$^{1}$Department of Physics and Astronomy, Bates College, Lewiston, ME USA\\
$^{2}$Company Name, Whereever, USA}

@@

We non-dimensionalise lengths in terms of the gap width $d = R_2 - R_1$ and velocities in terms of the inner cylinder velocity $U_i = R_i \Omega_i$. Taylor-Couette flow can be parameterised in terms of four parameters, the inner cylinder Reynolds number $\Reyn_i$, the ratio of inner to outer radii $\eta \equiv R_i/R_o$, the ratio of outer to inner rotation rates $\mu \equiv \Omega_o/\Omega_i$, and the aspect ratio of the cylinder $\Gamma \equiv L_z/d$. In our computations, the $z$ direction is periodic and thus $L_z$ is the periodicity length. We will also refer to the two Reynolds numbers $\Reyn_i$, $\Reyn_2o$, related by
\begin{equation}
    \frac{\Reyn_2}{\Reyn_1} = \frac{\mu}{\eta}.
\end{equation}
It is also quite useful to characterize times in terms of the viscous diffusion time $\tau_\nu \equiv d^2/\nu$. In our units, $\tau_\nu = \Reyn_i$.
We use the Dedalus framework \cite{2020PhRvR...2b3068B} for all linear and non-linear calculations. Pseudospectra and critical parameters are calculated using eigentools \cite{2021JOSS....6.3079O}.
\section{Linear Stability}
\label{sec:linear}
\subsection{Non-Axisymmetric Instability with $\mu = 0$}
%HAHAH This is all laid out in Krueger, Gross, Di Prima (1966), p 523-524
%The axisymmetric test case of \cite{1984JFM...146...65M} has $\mu = 0$, $\eta = 0.875$, $\Reyn = 139.31064$, and %$k_{min} = 2\pi/L_z = 2.513$ ($L_z = 2.5$). Interestingly, table~\ref{tab:critical_m} shows that circular Couette %flow is linearly unstable to all $m \leq 4$ modes at this $\Reyn$. 
%
%A natural assumption is that one would not expect to see axisymmetric flow at these parameters. Fully non-linear %simulations initialized from very low-amplitude random, $\mathbf{\nabla \cdot u} = 0$ perturbations bear this out. %Yet, \cite{1984JFM...146...65M} makes clear reference to TVF being stable to perturbations at this flow. He also %claims that initial conditions including ``round-off noise of the computer'' lead to axisymmetric flow. This is %quite unlikely to be true. 
%
%% the two runs (axisymmetric TVF and perturbed, restarted run) are on Leavitt
%One is naturally led to the question of how axisymmetric TVF is seen in experiment if $\Reyn_{c,m=1}/\Reyn_{c, %m=0} \simeq 1.008$: $m=1$ modes are unstable at a supercriticality of less than 1\%! 
%The answer becomes obvious if one considers the usual experimental procedure: spin up a fluid from rest to a %$\Reyn < \Reyn_{c,m=0}$, allow Couette flow to be established, then spin up until the first bifurcation occurs. %Allow the TVF to settle and proceed upwards. This will indeed produce an axisymmetric TVF flow that is %\emph{itself stable against $m\ne 0$ perturbations}! At $\Reyn = 138.31$, axisymmetric TVF flow can be perturbed %with gaussian noise and will remain axisymmetric. 
%
%\begin{table}[!h]
%\caption{Critical inner $\Reyn_i$ and $k_z$ as a function of $m$ for $\eta = 0.875$, $\mu = 0$}%%%Table caption %goes here
%\label{tab:critical_m}
%\begin{tabular}{rll}%%%The number of columns has to be defined here
%\hline
%m & $\Reyn_{ic}$ & $k_z$ \\
%\hline
%0 & 118.1 & 3.12947\\
%1 & 119.0 & 3.14374\\
%2 & 121.5 & 3.18888\\
%3 & 126.17 & 3.25852\\
%4 & 133.32 & 3.35672\\
%5 & 144.20 & 3.47885\\
%\end{tabular}
%\vspace*{-4pt}
%\end{table}%%%End of the table

@@

\subsection{$\Reyn_2 = -1000$, $\eta = 0.883$: Non-modal growth and GQL}
Next, we look at the case of $\Reyn_2 = -1000$. Setting $\eta = 0.883$ as in Andereck et al (1986), we consider the performance of GQL in both the linearly stable and unstable regime
\begin{itemize}
    \item look at pseudospectra for $\Reyn_1 = 450$, $\Reyn_1 = 400$ (below and above critical)
    \item compare pseudospectra between $\Reyn_2 = 0$ and $\Reyn_2 = -1000$
    \item do same thing at smaller $\Reyn_2$ to compare with \cite{2002PhFl...14.3475H}
    \item Compare GQL performance at various cutoffs
    \item Identify non-normality parameter for TC $\Lop$ for both linearly unstable state and non-linearly modified background
\end{itemize}
\subsection{Spiral Turbulence}
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{../figs/reo_rei_lsb.pdf}
    \caption{A summary of spiral turbulence runs in the $\Reyn_o-\Reyn_i$ plane showing linear stability boundary. Solid circles shows points where we have conducted GQL analysis; crosses represent DNS steps along the path to subcritical behaviour. }
    \label{fig:LSB}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figs/rei900_reo_-3398_KE_vs_t.pdf}
    \includegraphics[width=0.5\textwidth]{}
    \caption{Perturbation kinetic energy as a function of time for $\Reyn_i = 900, \Reyn_o=-3398$. Starting from a saturated spiral turbulence state at $t/\tau_\nu \simeq 1.5$, we continue the solution with QL ($\Lambda = 0$) and GQL with cutoffs $\Lambda = 1,3,5,10$. Note that all of the reduced models begin high amplitude oscillations.}
    \label{fig:ke_vs_t_rei900}
\end{figure}
%Meseguer et al (2009 PRE): 
%\begin{itemize}
%    \item $\Reyn_o = -1200, \Reyn_i = 600$, $\eta=0.883$, $\Gamma = 29.9$
%    \item $(N_r, N_\theta, N_z) = (20,220,220)$ HAHAHA. No way...more like $(64, 512, 512)$
%\end{itemize}

The spiral turbulence regime provides a unique opportunity to test GQL in an environment with both spatiotemporal patterns and a tunable bifurcation: by choosing $\Reyn_i$ and $\Reyn_o$ appropriately, both subcritical and supercritical routes to spiral turbulence can be selected.
Following pioneering numerical work \cite{2009PhRvE..79c6309M, 2009PhRvE..80d6315M}, we set the outer cylinder rotation rate to $\Reyn_o = -3398$ for the subcritical path and $\Reyn_o = -1359$ for the supercritical path and start a series of DNS runs at a $\Reyn_i$ high enough to trigger spiral turbulence driven by linear instability from a low-amplitude, random inital condition satisfying $\nabla \cdot \mathbf{u} = 0$.
From these two seed runs, we decrease $\Reyn_i$ in steps shown in figure~\ref{fig:LSB} where each point represents a DNS or GQL simulation run for one viscous time. The solid circles in figure~\ref{fig:LSB} represent runs where we performed GQL analyses.

\section{Runs}
\label{sec:nonlinear}

The ($\eta$, $\Gamma$) choice is motivated by \cite{1984JFM...146...45M,1984JFM...146...65M}, particularly section 4 of \cite{1984JFM...146...65M}.

@@

@@

\section{Resolution}
Might compare poorly resolved DNS to low order GQL as done in\cite{2017JFM...810..412T}. Note that in poorly resolved TC DNS, one gets spurious spiral modes.

@@

\appendix{Odds and ends}
\begin{itemize}
    \item On the $\Reyn_c(m=1)/\Reyn_c(m=0) \simeq 1.008$ issue...there's a discussion in  Krueger, Gross, Di Prima (1966), p 523-524, but it references a solution in a conference proceeding that is difficult to find. I have some memory of it being discussed in a more recent paper, but I can't find it anywhere...
\end{itemize}
\enlargethispage{20pt}

>>>>

\author{%%%% Author details
Jeffrey S. Oishi and Morgan Baxter}

%%%%%%%%% Insert author address here
\address{Department of Physics and Astronomy, Bates College, Lewiston, ME USA}

@@

We non-dimensionalise lengths in terms of the gap width $d = R_2 - R_1$ and velocities in terms of the inner cylinder velocity $U_i = R_i \Omega_i$. Taylor-Couette flow can be parameterised in terms of four parameters, the inner cylinder Reynolds number $\Reyn_i$, the ratio of inner to outer radii $\eta \equiv R_i/R_o$, the ratio of outer to inner rotation rates $\mu \equiv \Omega_o/\Omega_i$, and the aspect ratio of the cylinder $\Gamma \equiv L_z/d$. In our computations, the $z$ direction is periodic and thus $L_z$ is the periodicity length. We will also refer to the two Reynolds numbers $\Reyn_i$, $\Reyn_2o$, related by
\begin{equation}
    \frac{\Reyn_2}{\Reyn_1} = \frac{\mu}{\eta}.
\end{equation}
It is also quite useful to characterize times in terms of the viscous diffusion time $\tau_\nu \equiv d^2/\nu$. In our units, $\tau_\nu = \Reyn_i$.
We use the Dedalus framework (version 2) \cite{2020PhRvR...2b3068B} for all linear and non-linear calculations. Pseudospectra and critical parameters are calculated using eigentools \cite{2021JOSS....6.3079O}.
\section{Linear Stability}
\label{sec:linear}
\subsection{Non-Axisymmetric Instability with $\mu = 0$}
%HAHAH This is all laid out in Krueger, Gross, Di Prima (1966), p 523-524
%The axisymmetric test case of \cite{1984JFM...146...65M} has $\mu = 0$, $\eta = 0.875$, $\Reyn = 139.31064$, and %$k_{min} = 2\pi/L_z = 2.513$ ($L_z = 2.5$). Interestingly, table~\ref{tab:critical_m} shows that circular Couette %flow is linearly unstable to all $m \leq 4$ modes at this $\Reyn$. 
%
%A natural assumption is that one would not expect to see axisymmetric flow at these parameters. Fully non-linear %simulations initialized from very low-amplitude random, $\mathbf{\nabla \cdot u} = 0$ perturbations bear this out. %Yet, \cite{1984JFM...146...65M} makes clear reference to TVF being stable to perturbations at this flow. He also %claims that initial conditions including ``round-off noise of the computer'' lead to axisymmetric flow. This is %quite unlikely to be true. 
%
%% the two runs (axisymmetric TVF and perturbed, restarted run) are on Leavitt
%One is naturally led to the question of how axisymmetric TVF is seen in experiment if $\Reyn_{c,m=1}/\Reyn_{c, %m=0} \simeq 1.008$: $m=1$ modes are unstable at a supercriticality of less than 1\%! 
%The answer becomes obvious if one considers the usual experimental procedure: spin up a fluid from rest to a %$\Reyn < \Reyn_{c,m=0}$, allow Couette flow to be established, then spin up until the first bifurcation occurs. %Allow the TVF to settle and proceed upwards. This will indeed produce an axisymmetric TVF flow that is %\emph{itself stable against $m\ne 0$ perturbations}! At $\Reyn = 138.31$, axisymmetric TVF flow can be perturbed %with gaussian noise and will remain axisymmetric. 
%
%\begin{table}[!h]
%\caption{Critical inner $\Reyn_i$ and $k_z$ as a function of $m$ for $\eta = 0.875$, $\mu = 0$}%%%Table caption %goes here
%\label{tab:critical_m}
%\begin{tabular}{rll}%%%The number of columns has to be defined here
%\hline
%m & $\Reyn_{ic}$ & $k_z$ \\
%\hline
%0 & 118.1 & 3.12947\\
%1 & 119.0 & 3.14374\\
%2 & 121.5 & 3.18888\\
%3 & 126.17 & 3.25852\\
%4 & 133.32 & 3.35672\\
%5 & 144.20 & 3.47885\\
%\end{tabular}
%\vspace*{-4pt}
%\end{table}%%%End of the table

@@

%\subsection{$\Reyn_2 = -1000$, $\eta = 0.883$: Non-modal growth and GQL}
%Next, we look at the case of $\Reyn_2 = -1000$. Setting $\eta = 0.883$ as in Andereck et al (1986), we consider the performance of GQL in both the linearly stable and unstable regime
%\begin{itemize}
%    \item look at pseudospectra for $\Reyn_1 = 450$, $\Reyn_1 = 400$ (below and above critical)
%    \item compare pseudospectra between $\Reyn_2 = 0$ and $\Reyn_2 = -1000$
%    \item do same thing at smaller $\Reyn_2$ to compare with \cite{2002PhFl...14.3475H}
%    \item Compare GQL performance at various cutoffs
%    \item Identify non-normality parameter for TC $\Lop$ for both linearly unstable state and non-linearly modified background
%\end{itemize}
\subsection{Spiral Turbulence}
\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{../figs/reo_rei_lsb.pdf}
    \caption{A summary of spiral turbulence runs in the $\Reyn_o-\Reyn_i$ plane showing linear stability boundary. Solid circles shows points where we have conducted GQL analysis; crosses represent DNS steps along the path to subcritical behaviour. }
    \label{fig:LSB}
\end{figure}

%Meseguer et al (2009 PRE): 
%\begin{itemize}
%    \item $\Reyn_o = -1200, \Reyn_i = 600$, $\eta=0.883$, $\Gamma = 29.9$
%    \item $(N_r, N_\theta, N_z) = (20,220,220)$ HAHAHA. No way...more like $(64, 512, 512)$
%\end{itemize}

The spiral turbulence regime provides an ideal opportunity to test GQL in an environment with both spatiotemporal patterns and a tunable bifurcation: by choosing $\Reyn_i$ and $\Reyn_o$ appropriately, both subcritical and supercritical manifestations of spiral turbulence can be selected.
Following pioneering numerical work \cite{2009PhRvE..79c6309M, 2009PhRvE..80d6315M}, we set the outer cylinder rotation rate to $\Reyn_o = -3398$ for the subcritical path and $\Reyn_o = -1359$ for the supercritical path and start a series of DNS runs at a $\Reyn_i$ high enough to trigger spiral turbulence driven by linear instability from a low-amplitude, random inital condition satisfying $\nabla \cdot \mathbf{u} = 0$.
From these two seed runs, we decrease $\Reyn_i$ in steps shown in figure~\ref{fig:LSB} where each point represents a DNS or GQL simulation run for one viscous time. The solid circles in figure~\ref{fig:LSB} represent runs where we performed GQL analyses.

\section{Results}
\label{sec:nonlinear}
%The ($\eta$, $\Gamma$) choice is motivated by \cite{1984JFM...146...45M,1984JFM...146...65M}, particularly section 4 of \cite{1984JFM...146...65M}.

@@

\begin{figure}
    \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/rei900_reo_-3398_KE_vs_t.pdf}
    \end{minipage}
    \begin{minipage}{0.45\textwidth}
        \centering
        \includegraphics[width=0.9\textwidth]{figs/rei700_reo_-3398_KE_vs_t.pdf}
    \end{minipage}
    \caption{Perturbation kinetic energy as a function of time for $\Reyn_o=-3398$. On the left,  $\Reyn_i = 900$ is above the linear stability boundary, while the right shows $\Reyn_i = 700$, below the stability boundary. Starting from a saturated spiral turbulence state at $t/\tau_\nu \simeq 1.5$ (left), $t/\tau_\nu \simeq 3$ (right), we continue the solution with QL ($\Lambda = 0$) and GQL with cutoffs $\Lambda = 1,3,5,10$. For $\Reyn_i = 700$, we also use $\Lambda = 20, 30$ At higher $\Reyn_i$, all of the reduced models begin high amplitude oscillations. }
    \label{fig:ke_vs_t_rei900}
\end{figure}

\begin{figure}
    \centering
    \includegraphics{figs}
    \caption{Caption}
    \label{fig:my_label}
\end{figure}

@@

%\section{Resolution}
%Might compare poorly resolved DNS to low order GQL as done in\cite{2017JFM...810..412T}. Note that in poorly resolved TC DNS, one gets spurious spiral modes.

@@

%\appendix{Odds and ends}
%\begin{itemize}
%    \item On the $\Reyn_c(m=1)/\Reyn_c(m=0) %\simeq 1.008$ issue...there's a discussion in  Krueger, Gross, Di Prima (1966), p 523-524, but it references a solution in a conference proceeding that%is difficult to find. I have some memory of it being discussed in a more recent paper, but I can't find it anywhere...
%\end{itemize}
%\enlargethispage{20pt}
